---
title: "MotorTrend's Analysis of the Effects of Car Characteristics on MPG"
author: "S. Duffy"
date: "April 14, 2016"
output: word_document
---

##Summary



##Overview
Historically, the anecdote most car drivers pass down to their children is that a vehicle 
with a standard transmission is more fuel efficient than a vehicle with an automatic transmission.
We here at MotorTrend Magazine love and trust our dads but we wanted to see if this anecdote 
stands up to an analysis of the data.

We were able to obtain rare data from industry insiders and have stored it in a dataset called
mtcars.  The dataset has a structure like this:

```{r}
str(mtcars)
```

The definitions of the columns are thus:

* mpg	= Miles/(US) gallon
* cyl	= Number of cylinders
* disp	= Displacement (cu.in.)
* hp	= Gross horsepower
* drat	= Rear axle ratio
* wt	= Weight (1000 lbs)
* qsec	= 1/4 mile time
* vs	= V/S
* am	= Transmission (0 = automatic, 1 = manual)
* gear	= Number of forward gears
* carb	= Number of carburetors

What we're interested in is the mpg and am columns which have the following characteristics

```{r}
a <- summary(mtcars$mpg)
b <- nrow(mtcars)
man <- sum(mtcars$am)
auto <- b-man
```

The mpg column has a range from `r a[1]` mpg to `r a[6]` mpg with a mean of `r a[4]`.  

Out of the `r b` vehicles in the dataset, `r man` are manual transmissions and `r auto` are automatic transmissions.

Is there are noticeable difference in MPG in the two different transmission types?  A plot shows yes.
The blue line is the average for 1s, the manual transmissions and the red line is the average for 
0s, the automatic transmissions.

```{r}
mancars <- subset(mtcars, am == 0)
autocars <- subset(mtcars, am == 1)

plot(mtcars$am,mtcars$mpg, ylab = "MPG", xlab = "Transmission 1=auto", main = "MPG by Transmission with Avgs")
  abline(h=mean(mancars$mpg),col = 'red')
  abline(h=mean(autocars$mpg),col = 'blue')
```


##Is an automatic or manual transmission better for MPG?
The question we'll explore is, "Is an automatic or manual transmission better for MPG?"  Of course, 
there are a lot of moving pieces in a vehicle that can contribute to varying levels of MPG, many of 
which are in the dataset.  First, we'll explore our question just using the two variables in the 
question, mpg and am.

For this section, we'll use the mpg as the dependent variable and transmission as a predictor

```{r}
mdl <- lm(mpg ~ am,mtcars)
summary(mdl)  
```

Much like in the plot above, one can see the intercept is at the average of the automatic transmission
vehicles and the slope rises 7.25 MPG to the manual transmission. The coeffiecient of the slope for auto/
manual has a p value of 0.000285, thus, we can reject the null hypothesis that there is no difference
in the mpg between manual and automatic transmissions.


##

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
